---
title: "Walsh - Assignment 1"
author: "Sharon Walsh"
date: "2024-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Walsh Assignment 1

This dataset was downloaded from Kaggle.com. This data is called "QS World Ranked Universities (2018-2022). The link is as follows: https://www.kaggle.com/datasets/aklimarimi/qs-world-ranked-universities-20182022



```{r}
#install.packages("caret")
library(caret)
#install.packages("ISLR")
library(ISLR)
setwd("C:/Users/sharo/OneDrive/Documents/ML 2024/Assignment 1")
DF<-read.csv("Walsh Data - Assignment 1.csv")
summary(DF)
```

## Descriptive Statistics 

## Quantitative variables

```{r}
cat("The mean of Rank is", mean(DF$Rank))
```

```{r}
cat("The median of Rank is", median(DF$Rank))
```

```{r}
cat("The IQR of Rank is", print(IQR(DF$Rank)))
```
```{r}
cat("The standard deviation of Rank is", print(sd(DF$Rank)))
```

```{r}
cat("The correlation between Rank and Year is", cor(DF$Rank, DF$Year))
```
##Descriptive Statistics

## Categorical Variables

```{r}
library(dplyr)
DF%>%group_by(Country)%>%summarise(City=n())%>%as.data.frame()
```


```{r}
Table1 <- table(DF$City)
head(Table1)
```
##Transform 1 variable

## Transform Variable

```{r}
Table2 <- as.character(DF$Year)
head(Table2)
```
##Plot 1 quantitative variable

```{r}
head(DF)
hist(DF$Point)
```

This historgram show that the "Point" data follows a normal distribution. 

##1 scatterplot

```{r}
plot(DF$Rank)
```
```{r}
library(ggplot2)
ggplot(DF, aes(x=Point, y=Rank))+geom_point()
```

